One-step, three-factor authentication with in-ear EEG

### 2016-12-01 paper.tex
added data points for classifier

% Created 2016-11-30 Wed 17:09
% Intended LaTeX compiler: pdflatex
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\author{Max Curran and Nick Merrill}
\date{\today}
\title{One-step, three-factor authentication with in-ear EEG}
\hypersetup{
 pdfauthor={Max Curran and Nick Merrill},
 pdftitle={One-step, three-factor authentication with in-ear EEG},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 26.0.50.2 (Org mode 9.0.1)}, 
 pdflang={English}}
\begin{document}

\maketitle
\begin{abstract}
We propose the first research study of one-step three-factor authentication.
Specifically we seek to demonstrate its feasibility and quantify its performance
using Ear EEG (electroencephalogram). By performing a single mental task, our
user will be able to present three authenticators at once – a knowledge factor
(their chosen secret thought and/or mental task), an inherence factor (their
brainwave signals), and a possession factor (the EEG sensing earpiece that is
custom-fitted to their ear). We will build the custom-fit Ear EEG earpieces and
conduct an experimental study to evaluate the accuracy and usability of this
authentication method.
\end{abstract}
\section{Introduction and Motivation}
\label{sec:orgd88076b}

@@

\section{Related Work}
\label{sec:orge0e46dc}
The use of EEG as a biometric signal for user authentication has a short history.
In 2005, Thorpe et al. motivate and outline the design of a passthoughts system,
where users authenticate by thinking of a passthought rather than typing in a
password [6]. Since 2002, a number of independent groups have achieved 99-
100\% authentication accuracy using multi-channel sensors placed on the scalp
[7-10]. In 2013, our group showed that 99\% authentication accuracy can also be
achieved using a consumer-grade single-channel sensor [2]. In particular, the
lack of signal diversity from multiple EEG channels can be overcome by allowing
the users to choose their own personalized passthoughts (e.g., sing their favorite
song in their head). There are two significant consequences of this result. First,
the passthoughts approach is no longer constrained by the high cost (\textasciitilde{}\$10k’s)
and low usability (gel-based electrodes; aesthetic challenges of an EEG cap) of
medical-grade multi-channel devices. Second, because users can choose and
easily change their secret mental task, this approach can support one-step two-
factor authentication [1] via the simultaneous presentation of the inherence factor
(brainwave signatures due to the unique folding structures of the cortex) and the
knowledge factor (the secret mental task).

@@

\section{Proposed project}
\label{sec:org6246122}

@@

\section{Research plan}
\label{sec:org5e36c1a}

@@

For authentication analysis, we will build upon the analysis methodology and
tools from our previous studies. We will apply the logarithmic binning technique
that we have developed in [5] to the EEG power spectrum data during the signal
pre-processing step. We will then apply machine learning to train and test a
support vector classifier as in [4]. We will compare the results against the
threshold-based authentication protocol based on cosine similarity as in [2,3].
A key change for this study is that we will be able to collect up to 8 separate
channels of EEG data, including multiple channels from each ear. This is in
contrast to our previous studies, where we collected only a single channel of
EEG, either on the prefrontal cortex (FP1) location or in the ear canal. Therefore,
we will need to develop new analytic techniques to both (i) assess improvements
in authentication accuracy due to the increased channels of data, and (ii) assess
relative contributions of individual electrode locations to authentication accuracy.
Finally, we will evaluate the usability of custom-fit Ear EEG authentication along
several dimensions. They include the comfort and fit of the earpieces, the
preparatory steps of using the devices, the ease and repeatability of the
authentication tasks, and the recall rates of the personal secrets.
\section{{\bfseries\sffamily TODO} Preliminary results}
\label{sec:orgb471f1b}

@@

\subsection{{\bfseries\sffamily TODO} Validation}
\label{sec:orgdae54ad}
\subsection{{\bfseries\sffamily TODO} Authentication performance}
\label{sec:org14de4b9}
\subsection{Disentangling inherence and knowledge}
\label{sec:org91c37b4}

@@

\begin{center}
\begin{tabular}{lrrr}
Task & Right ear & Left ear & fp1\\
\hline
speech & 0.311 & 0.116 & 0\\
face & 0.355 & 0.055 & 0\\
breathe & 0 & 0.240 & 0\\
listennoise & 0.322 & 0.078 & 0\\
song\(_{\text{o}}\) & 0 & 0.166 & 0\\
song & 0 & 0.211 & 0\\
sequence & 0.433 & 0 & 0\\
listentone & 0 & 0.333 & 0\\
breathe\(_{\text{o}}\) & 0.254 & 0.129 & 0\\
sport & 0.222 & 0.135 & 0\\
\hline
\textbf{p-value} & \textbf{*0.0076} & \textbf{*0.009} & N/A\\
\end{tabular}
\end{center}
Our low p-values (Table 3) indicate that the passthought, as well as the inherence
factor associated with the custom-fit EEG earbud, both contribute to the classifier
performance discussed in Section 5.2. We are confident, then, in our claim that this work
could potentially yield multiple-factor authentication in a single step. However, we
will need to collect a great deal more data to substantiate this claim rigorously.

\section{Schedule and Budget}
\label{sec:org1590dbf}
We propose a 12-month schedule with two graduate students working on the
project:

\begin{itemize}
\item Months 1-6: hardware development for custom-fit earpiece; integration with OpenBCI; testing
\item Months 1-6: software integration between OpenBCI, PsychoPy, and Indra; design of new authentication tasks and implementation in PsychoPy
\item Months 7-8: experimental data collection
\item Months 9-12: data analysis; write and submit research papers
\end{itemize}
\section{References}
\label{sec:org14ba974}

>>>>

% Created 2016-11-30 Wed 17:24
% Intended LaTeX compiler: pdflatex
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\author{Max Curran and Nick Merrill}
\date{\today}
\title{One-step, three-factor authentication with in-ear EEG}
\hypersetup{
 pdfauthor={Max Curran and Nick Merrill},
 pdftitle={One-step, three-factor authentication with in-ear EEG},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 26.0.50.2 (Org mode 9.0.1)}, 
 pdflang={English}}
\begin{document}

\maketitle
\begin{abstract}
We propose the first research study of one-step three-factor authentication.
Specifically we seek to demonstrate its feasibility and quantify its performance
using Ear EEG (electroencephalogram). By performing a single mental task, our
user will be able to present three authenticators at once – a knowledge factor
(their chosen secret thought and/or mental task), an inherence factor (their
brainwave signals), and a possession factor (the EEG sensing earpiece that is
custom-fitted to their ear). We will build the custom-fit Ear EEG earpieces and
conduct an experimental study to evaluate the accuracy and usability of this
authentication method.
\end{abstract}
\section{Introduction and Motivation}
\label{sec:org557c212}

@@

\section{Related Work}
\label{sec:org08d7c07}
The use of EEG as a biometric signal for user authentication has a short history.
In 2005, Thorpe et al. motivate and outline the design of a passthoughts system,
where users authenticate by thinking of a passthought rather than typing in a
password [6]. Since 2002, a number of independent groups have achieved 99-
100\% authentication accuracy using multi-channel sensors placed on the scalp
[7-10]. In 2013, our group showed that 99\% authentication accuracy can also be
achieved using a consumer-grade single-channel sensor [2]. In particular, the
lack of signal diversity from multiple EEG channels can be overcome by allowing
the users to choose their own personalized passthoughts (e.g., sing their favorite
song in their head). There are two significant consequences of this result. First,
the passthoughts approach is no longer constrained by the high cost (\textasciitilde{}\$10k’s)
and low usability (gel-based electrodes; aesthetic challenges of an EEG cap) of
medical-grade multi-channel devices. Second, because users can choose and
easily change their secret mental task, this approach can support one-step two-
factor authentication [1] via the simultaneous presentation of the inherence factor
(brainwave signatures due to the unique folding structures of the cortex) and the
knowledge factor (the secret mental task).

@@

\section{Proposed project}
\label{sec:orge060800}

@@

\section{Research plan}
\label{sec:orgad6a419}

@@

For authentication analysis, we will build upon the analysis methodology and
tools from our previous studies. We will apply the logarithmic binning technique
that we have developed in [5] to the EEG power spectrum data during the signal
pre-processing step. We will then apply machine learning to train and test a
support vector classifier as in [4]. We will compare the results against the
threshold-based authentication protocol based on cosine similarity as in [2,3].
A key change for this study is that we will be able to collect up to 8 separate
channels of EEG data, including multiple channels from each ear. This is in
contrast to our previous studies, where we collected only a single channel of
EEG, either on the prefrontal cortex (FP1) location or in the ear canal. Therefore,
we will need to develop new analytic techniques to both (i) assess improvements
in authentication accuracy due to the increased channels of data, and (ii) assess
relative contributions of individual electrode locations to authentication accuracy.
Finally, we will evaluate the usability of custom-fit Ear EEG authentication along
several dimensions. They include the comfort and fit of the earpieces, the
preparatory steps of using the devices, the ease and repeatability of the
authentication tasks, and the recall rates of the personal secrets.
\section{{\bfseries\sffamily TODO} Preliminary results}
\label{sec:org9310aa7}

@@

\subsection{{\bfseries\sffamily TODO} Validation}
\label{sec:org9fa2fd6}
\subsection{{\bfseries\sffamily TODO} Authentication performance}
\label{sec:org993972c}

Following [5], we \ldots{}..

\begin{center}
\begin{tabular}{lrrrrrr}
task & Left ear &  & Right Ear &  & fp1 & \\
\hline
 & FAR & FRR & FAR & FRR & FAR & FRR\\
\hline
speech & 0 & 0 & 0 &  & 0 & 0\\
face & 0 & 0 & 0 & 0.002 & 0 & 0\\
breathe & 0 & 0 & 0 &  & 0 & 0\\
listennoise & 0 & 0 & 0 &  & 0 & 0\\
song (eyes open) & 0 & 0 & 0 &  & 0 & 0\\
song & 0 & 0 & 0 &  & 0 & 0\\
sequence & 0 & 0 & 0 &  & 0 & 0\\
listentone & 0 & 0 & 0 & 0.009 & 0 & 0\\
breathe (eyes open) & 0.039 & 0 & 0 &  & 0 & 0\\
sport & 0.039 & 0.036 & 0 &  & 0 & 0\\
\end{tabular}
\end{center}

\subsection{Disentangling inherence and knowledge}
\label{sec:orgec74458}

@@

\begin{center}
\begin{tabular}{lrrr}
Task & Right ear & Left ear & fp1\\
\hline
speech & 0.311 & 0.116 & 0\\
face & 0.355 & 0.055 & 0\\
breathe & 0 & 0.240 & 0\\
listennoise & 0.322 & 0.078 & 0\\
song\(_{\text{o}}\) & 0 & 0.166 & 0\\
song & 0 & 0.211 & 0\\
sequence & 0.433 & 0 & 0\\
listentone & 0 & 0.333 & 0\\
breathe\(_{\text{o}}\) & 0.254 & 0.129 & 0\\
sport & 0.222 & 0.135 & 0\\
\hline
\textbf{p-value} & \textbf{*0.0076} & \textbf{*0.009} & N/A\\
\end{tabular}
\end{center}
Our low p-values (Table 2) indicate that the passthought, as well as the inherence
factor associated with the custom-fit EEG earbud, both contribute to the classifier
performance discussed in Section 5.2. We are confident, then, in our claim that this work
could potentially yield multiple-factor authentication in a single step. However, we
will need to collect a great deal more data to substantiate this claim rigorously.

\section{Schedule and Budget}
\label{sec:org2f6eb19}
We propose a 12-month schedule with two graduate students working on the
project:

\begin{itemize}
\item Months 1-6: hardware development for custom-fit earpiece; integration with OpenBCI; testing
\item Months 1-6: software integration between OpenBCI, PsychoPy, and Indra; design of new authentication tasks and implementation in PsychoPy
\item Months 7-8: experimental data collection
\item Months 9-12: data analysis; write and submit research papers
\end{itemize}
\section{References}
\label{sec:org72eb964}
